Tensor anisotropy as a tracer of cosmic voids

### 2015-02-18 cosmic_voids.tex
clarifying text, cleaning typos

Here we introduce a new algorithm to define voids over the continous
matter density or velocity distribution defined on a fixed and
homogeneous spatial grid.
The algorithm is based on two tensorial schemes used to classify the
cosmic web. 
The first (the T-web) is based on the Hessian of the gravitational potential or
tidal tensor \citep{Hahn07, Forero09}. 
The second (the V-web) is based on the velocity shear tensor
\citep{Hoffman12}. 
Our procedure allows a description of the internal structure
of voids that goes beyond a simple definition of a void as just an
underdense regions in the large-scale matter distribution.
The tidal and the shear tensors encode more information than the
density field as they trace the collapsing or expanding
nature of the matter field, which defines the dynamics of the Cosmic
Web.

@@

Once we establish the FA as a void tracer, we proceed to
identify individual voids as basins of FA local minima. 
At this point we implement a \textit{watershed transform algorithm}
\citep{Beucher79,Beucher93} which has been used to define voids as
catching basins of local minima of the density field
\citep{Platen07,Neyrinck08}.

%This paper is organized as follows. 
%In Section \ref{sec:algorithms_cosmic_web} we describe the algorithms
%we use to find the cosmic web in N-body simulations. 
%In Section \ref{sec:bulk_voids} we explain in detail our void finder
%based on the fractional anisotropy of the tidal and shear tensor
%fields. 
%In Section  \ref{sec:simulations} we describe the N-body simulation we
%use to test our algorithm. 
%Our results are presented in \ref{sec:results} to finally comment and
%conclude about these findings in Section \ref{sec:conclusions}.

@@

%.........................................................................
%FIGURE 1: FA and vissual impression
\begin{figure*}
  \includegraphics[trim = 16mm 8mm 5mm 12mm, clip, keepaspectratio=true,
  width=0.73\textheight]{cosmicweb_FA_Tweb.pdf}
  \includegraphics[trim = 16mm 8mm 5mm 12mm, clip, keepaspectratio=true,
  width=0.73\textheight]{cosmicweb_FA_Vweb.pdf}
  \caption{Visual impression of different quantities/algorithm results
    mentioned in this paper. Fractional Anisotropy field
    (left); osmic web classification (middle) where voids are white,
    sheets are light gray, filaments are dark gray and knots are
    black; and colored individual voids. Top/bottom corresponds to the 
    T-Web/V-Web.}
  \label{fig:FA_field}
\end{figure*}
%.........................................................................

@@

This scheme was initially proposed by \citet{Hahn07} as an
alternative for classifying the Cosmic Web based on the tidal tensor.
The tidal tensor allows a classification in terms of the orbital
dinamics of the matter field.
This approach extends to second-order the equations of motion around 
local minima of the gravitational potential. 
The second-order term corresponds to the tidal tensor, which is
defined as the Hessian matrix of the normalized gravitational
potential.

@@

The fractional anisotropy (FA), as developed by \citet{Basser95}, was 
conceived to quantify the anisotropy degree of a diffusion process, e.g. 
the diffusivity of water molecules through cerebral tissue in nuclear
magnetic resonance imaging. 
Here we present the FA, much in the same way as \citet{Libeskind13},
to use it as a tracer of cosmic voids.

The FA is defined as follows.
%.........................................................................
%Fractional anisotropy
\eq{fractional_anisotropy}
{{\mathrm{FA}} = \frac{1}{\sqrt{3}}\sqrt{ \frac{ (\lambda_1 - \lambda_3)^2 + 
(\lambda_2 - \lambda_3)^2 + (\lambda_1 - \lambda_2)^2}{ \lambda_1^2 + 
\lambda_2^2 + \lambda_3^2} },}
%.........................................................................
where the eigenvalues can be taken from either the T-web or the Vweb 
(FA-T-web and FA-Vweb respectively). Such as it is defined, FA$=0$ 
corresponds to an isotropic distribution ($\lambda_1=\lambda_2=\lambda_3$) 
and FA$=1$ with a highly anisotropic distribution.

In the left and middle panels of Figure \ref{fig:FA_field} we show the
FA field and web classification for both  web schemes over a slice of
an N-body simulation (described in Section \ref{sec:simulations}). 
Comparing these two panels we see that voids and knots (white
and black in the middle panel of Figure 1) display low FA values at
their  centres, becoming gradually more anisotropic at outer regions.  
On the other hand the filamentary structure (grey in the middle panel
of Figure 1) is traced by high FA values.  
These characteristics are key to use the FA as a tracer of cosmic voids.

@@

  \captionof{figure}{\small Overdensity (left) and Fractional
    Anisotropy (right) with respect to the eigenvalue $\lambda_1$ 
    for each web scheme (T-web, continous lines. Vweb, dashed lines) calculated 
    over all cells of the grid. Thick central lines correspond are the
    median and filled regions include $50\%$ of all values. This
    information is necesary to define void boundaries in the watershed
    algorithm. The vertical lines in the right hand plot show the
    equivalent threshold values for the T-Web and V-web in order to
    have voids composed by cells with Fractional Anisotropy equal or
    less than $0.95$.}
  \label{fig:L1_correlations}

@@

From Figs. \ref{fig:FA_field} and \ref{fig:L1_correlations} we
conclude that the FA is a good tracer of voids as it is almost perfectly
correlated with low values of $\lambda_1$. 
We propose that voids should be composed completely by regions of
FA$<0.95$.
If we increase the values of $\lambda_1$ from its minimum until it
we reach FA$=0.95$ in \ref{fig:L1_correlations} we find that this
correspond to critical values of $\lambda_{1}^T = 0.265$ and
$\lambda_{1}^V = 0.175$ for the T-web and Vweb, respectively.
This means that setting $\lambda_{th}$ to either
$\lambda_{1}^T$/$\lambda_{1}^{V}$ automatically produces voids with
all the cells FA$<0.95$.   
The middle panels in Figure \ref{fig:FA_field} show the web
classification for this choice of $\lambda_{th}$, demonstrating that
this FA level is a sensitive choice to define voids.

%-------------------------------------------------------------------------
\subsection{Defining voids with a watershed algorithm}
\label{sec:watershed}
%-------------------------------------------------------------------------

The previous section shows that FA is a good void tracer, but it does
not automatically suggest how to define the boundary of individual
voids.
For this purpose, we use the \textit{watershed transform algorithm}
\citep{Beucher79,Beucher93} to identify a void as the basin of FA
local minimum. 
The advantage of this definition is that it does not require any
assumption on the shape and/or morphology of the tentative voids.

However, there are two main differences in our approach with respect to other
watershed implementations.
First, the watershed technique commonly uses the the density field
instead of the FA field as we do in this paper
\citep{Platen07,Neyrinck08}.   
Second, we estimate all relevant quantities on a Cartesian mesh of
fixed cell size, while other works use an adaptive Delaunay tessellation
\citep{Schaap00}.
However, from the analysis of our results it does not 
introduce supurious results, at least with the mass resolution we have
in the N-body simulation.

@@

In Figure \ref{fig:L1_correlations} we find the correlations of
the eigenvalue $\lambda_1$ with he matter overdensity (left panel) and
the FA (right panel).
We can use this information to find the FA threshold. 
From the left panel we see that an underdensity of $\delta=-0.8$ is
correspond to values of $\lambda_1=0.0$, regardless of the web-finding
scheme.
From the second panel we have a very tight correlation of $\lambda_1$
with the FA, indicating that in turn $\lambda_1=0.0$ corresponds to a
FA=$0.65$ which is the value that we have used to remove ridges.

The right column in Figure \ref{fig:FA_field} shows all the individual
voids that have been identified using the watershed algorithm on the
FA field. 
In what follows we describe the numerical simulation we have used to
find voids and their detailed properties characterization.

@@

We see that the void distribution for the T-Web is broadly consistent with the
expectations from a two-barrier problem  \citep{Sheth04}. 
The formation of large voids is limited by the \textit{void-in-void}
problem (first barrier), where large voids are constituted
hierarchically of smaller ones. 
In turn, the formation of small voids is damped by the
\textit{void-in-cloud} problem (second barrier),  where nearby
collapsing structures limit the abundance of small embedded voids.

We also find that the V-Web scheme produces an over-abundance of small
voids compared to the resuls of the T-Web.
A large number of these small voids are embedded in overdense regions. 
They are visible in the middle panel of Figure \ref{fig:FA_field} as
small bubbles located inside sheets.
The existence of these small voids can be explained by
dynamics of shell crossing in collapsing sheets. 
As matter colides into a sheet, in their symmetry plane one will
find crossing ahseets that effectively give a positive divergence in
the velocity field, resulting in a void identification by the
V-Web algorithm. This points has been discussed in \citet{Hoffman12}.

Comparing the abundance of large voids in the two web schemes, we find
that the V-Web is limited to have voids on the scale of $\sim 10$\hMpc,
while the T-Web scheme includes voids as large as $\sim 15$\hMpc. 
Large voids in the T-Web scheme have a velocity structure that
induces a split by the watershed algorithm in the V-Web. 
This is evident in the right panel of Figure \ref{fig:FA_field}.

>>>>

Here we introduce a new algorithm to define voids over the continous
matter density or velocity distribution defined on a fixed and
homogeneous spatial grid.
The algorithm uses the results from two tensorial schemes used to
classify the cosmic web.
The first (the T-web) is based on the Hessian of the gravitational potential or
tidal tensor \citep{Hahn07, Forero09}. 
The second (the V-web) is based on the velocity shear tensor
\citep{Hoffman12}. 
Our procedure allows for a description of the internal structure
of voids that goes beyond a simple definition of a void as an
underdense region in the large-scale matter distribution.
The tidal and the shear tensors encode more information than the
density field as they trace the collapsing or expanding
nature of the matter field, which defines the dynamics of the Cosmic
Web.

@@

In the next sections (\S \ref{sec:algorithms_cosmic_web} and \S
\ref{sec:bulk_voids}) we establish the FA as a void tracer and then
proceed to identify individual voids as basins of FA local minima.  At
this point we implement a \textit{watershed transform algorithm}
\citep{Beucher79,Beucher93} which has been used to define voids as
catching basins of local minima of the density field \citep{Platen07,Neyrinck08}.
Finally we find and charachterize voids in a N-body (\S
\ref{sec:simulations} and \S \ref{results}) using density, velocity and
fractional anisotropy profiles.  We use these results to comment on
the qualities of our algorithm (\S \ref{sec:conclusions}).

@@

%.........................................................................
%FIGURE 1: FA and vissual impression
\begin{figure*}
  \includegraphics[trim = 16mm 8mm 5mm 12mm, clip, keepaspectratio=true,
  width=0.73\textheight]{cosmicweb_FA_Tweb.pdf}
  \includegraphics[trim = 16mm 8mm 5mm 12mm, clip, keepaspectratio=true,
  width=0.73\textheight]{cosmicweb_FA_Vweb.pdf}
  \caption{Visual impression of different quantities/algorithm results
    mentioned in this paper. Fractional Anisotropy field
    (left); cosmic web classification (middle) where voids are white,
    sheets are light gray, filaments are dark gray and knots are
    black; and colored individual voids. Top/bottom corresponds to the 
    T-Web/V-Web.}
  \label{fig:FA_field}
\end{figure*}
%.........................................................................

@@

This scheme was initially proposed by \citet{Hahn07}  to find the
Cosmic Web based on the tidal tensor. 
The tidal tensor allows a classification in terms of the orbital
dinamics of the matter field.
This approach extends to second-order the equations of motion around 
local minima of the gravitational potential. 
The second-order term corresponds to the tidal tensor, which is
defined as the Hessian matrix of the normalized gravitational
potential.

@@

The fractional anisotropy (FA), as developed by \citet{Basser95}, was 
conceived to quantify the anisotropy of the diffusivity of water
molecules through cerebral tissue in nuclear magnetic resonance imaging. 
Here we present the FA, much in the same way as \citet{Libeskind13},
to use it as a tracer of cosmic voids.

The FA is defined as follows.
%.........................................................................
%Fractional anisotropy
\eq{fractional_anisotropy}
{{\mathrm{FA}} = \frac{1}{\sqrt{3}}\sqrt{ \frac{ (\lambda_1 - \lambda_3)^2 + 
(\lambda_2 - \lambda_3)^2 + (\lambda_1 - \lambda_2)^2}{ \lambda_1^2 + 
\lambda_2^2 + \lambda_3^2} },}
%.........................................................................
where the eigenvalues can be taken from either the T-Web or the V-Web 
(FA-T-Web and FA-V-Web respectively). Such as it is defined, FA$=0$ 
corresponds to an isotropic distribution ($\lambda_1=\lambda_2=\lambda_3$) 
and FA$=1$ to a highly anisotropic distribution.

In the left and middle panels of Figure \ref{fig:FA_field} we show the
FA field and web classification for both  web schemes over a slice of
an N-body simulation (described in Section \ref{sec:simulations}). 
Comparing these two panels we see that voids and knots (white
and black in the middle panel of Figure 1) display low FA values at
their  centres, becoming gradually more anisotropic at outer regions.  
On the other hand the filamentary structure (grey in the middle panel
of Figure 1) is traced by high FA values.  
This characteristic is the  key to use the FA as a tracer of cosmic
voids.

@@

  \captionof{figure}{\small Overdensity (left) and Fractional
    Anisotropy (right) with respect to the eigenvalue $\lambda_1$ 
    for each web scheme (T-Web, continous lines. V-Web, dashed lines) calculated 
    over all cells of the grid. Thick central lines correspond are the
    median and filled regions include $50\%$ of all values. This
    information is necesary to define void boundaries in the watershed
    algorithm. The vertical lines in the right hand plot show the
    equivalent threshold values for the T-Web and V-web in order to
    have voids composed by cells with Fractional Anisotropy equal or
    less than $0.95$.}
  \label{fig:L1_correlations}

@@

From Figs. \ref{fig:FA_field} and \ref{fig:L1_correlations} we
conclude that the FA is a good tracer of voids as it is almost perfectly
correlated with low values of $\lambda_1$. 
We propose that voids should be composed completely by regions of
FA$<0.95$.
If we increase the values of $\lambda_1$ from its minimum until it
we reach FA$=0.95$ in \ref{fig:L1_correlations} we find that this
correspond to critical values of $\lambda_{1}^T = 0.265$ and
$\lambda_{1}^V = 0.175$ for the T-Web and V-Web, respectively.
This means that setting $\lambda_{th}$ to either
$\lambda_{1}^T$/$\lambda_{1}^{V}$ automatically produces voids with
all the cells FA$<0.95$.   
The middle panels in Figure \ref{fig:FA_field} show the web
classification for this choice of $\lambda_{th}$, demonstrating that
this FA level is a sensitive choice to define voids.

%-------------------------------------------------------------------------
\subsection{Defining voids with a watershed algorithm}
\label{subsec:watershed}
%-------------------------------------------------------------------------

The previous Section shows that FA is a good void tracer, but how
should we actually define the boundary of individual voids?
For this purpose, we use the \textit{watershed transform algorithm}
\citep{Beucher79,Beucher93} to identify a void as the basin of FA
local minima with boundaries of FA$=0.95$. 
The advantage of this definition is that it does not require any
assumption on the shape and/or morphology of the tentative voids.

However, there are two main differences in our approach with respect to other
watershed implementations.
First, the watershed technique commonly uses the density field
instead of the FA field as we do here
\citep{Platen07,Neyrinck08}.   
Second, we estimate all relevant quantities on a Cartesian mesh of
fixed cell size, while other works use an adaptive Delaunay tessellation
\citep{Schaap00}.
However, from the analysis of our results the choice of a fixed
Cartesian mesh does not induce wrong results, at least with
the mass resolution we have in the N-body simulation.

@@

In Figure \ref{fig:L1_correlations} we find the correlations of
the eigenvalue $\lambda_1$ with the matter overdensity (left panel) and
the FA (right panel).
We can use this information to find the equivalent FA threshold. 
From the left panel we see that an underdensity of $\delta=-0.8$ 
corresponds to $\lambda_1=0.0$, regardless of the web-finding
scheme.
From the second panel we have a very tight correlation of $\lambda_1$
with the FA, indicating that in turn $\lambda_1=0.0$ corresponds to a
FA=$0.65$ which is the value that we choose to remove ridges. 
The right column in Figure \ref{fig:FA_field} shows all the individual
voids that have been identified using the watershed algorithm on the
FA field.

@@

We see that the void distribution for the T-Web is broadly consistent with the
expectations from a two-barrier problem  \citep{Sheth04}. 
The formation of large voids is limited by the \textit{void-in-void}
mechanism (first barrier), where large voids are constituted
hierarchically of smaller ones. 
In turn, the formation of small voids is damped by the
\textit{void-in-cloud} mechanism (second barrier),  where nearby
collapsing structures limit the abundance of small embedded voids.

We also find that the V-Web scheme produces an over-abundance of small
voids compared to the resuls of the T-Web.
A large number of these small voids are embedded in overdense regions. 
They are visible in the middle panel of Figure \ref{fig:FA_field} as
tiny white bubbles located inside sheets.
The existence of these small voids can be explained by
dynamics of shell crossing in collapsing sheets. 
As matter colides into a sheet, in their symmetry plane one will
find crossing sheets that effectively give a positive divergence in
the velocity field, resulting in a void identification by the
V-Web algorithm. This has been already discussed in \citet{Hoffman12}.

Comparing the abundance of large voids in the two web schemes, we find
that the V-Web is limited to have voids on the scale of $\sim 10$\hMpc,
while the T-Web scheme includes voids as large as $\sim 15$\hMpc. 
Large voids in the T-Web scheme have a velocity structure that
induces a split by the watershed algorithm in the V-Web. 
This increased granularity in the velocity structure of voids is
evident in the left and right panels of Figure \ref{fig:FA_field}.
